Notes from the President

### 2022-04-08 pnsac/newsletter/article_Feature1.tex
First set of changes for 18-1 issue, Apr.2022.

\title{North Star Propeller Restorations}
%\subtitle{Part 4}
\author{Harry Hope, with assistance from Michael Hope.}

@@

\textit{Hope Aero has played a major role in the restoration of the North Star.
The following article provides some background about the company and its work
on the North Star’s propellers. }\\

Hope Aero was formally government approved to certify aircraft propellers in
1992, at their facility in Mississauga, Ontario. Harry Hope however, already
had 40 years of propeller overhaul experience on many varieties that included
Hamilton Standard propellers, which were quite common, as well as Hartzell,
McCauley, and Sensenich. Harry's first restoration to flying condition, was the
propeller for the Lysander, which was restored by the Air Force in Winnipeg for
their Centennial project in 1967. This Lysander is now displayed at CASM. Since
that time, they have restored several propellers, some to flying condition,
three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
display at the Aviation museum in Boda, Norway.

 Since that time, their capabilities have added– MT, Dowty, Hoffman, and
Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
their customer base expanded to include different types of aircraft, the
requirement for them to service more types became evident. They kept up with
the required technology.

Along the way, they also expanded into aircraft Wheel and Brake overhaul,
sales, service and on wing dynamic balancing of fixed wing and Helicopters.
Harry retired in 2001, but the third generation is now capably looking after
the business.

In the spring of 2006, Hope Aero took on the task of restoring the four North
Star propellers, for the "North Star Restoration" project. The first challenge
was to remove the propellers from the engines as neither the restoration crew
nor Hope Aero had any recent experience removing this type of propeller.
Michael Hope made the trip to Ottawa to assist with this operation and to help
prevent any damage, which turned out to be exceedingly difficult due to the
corrosion, starting with the spinner and then the propeller. Once the propeller
was removed the blades were removed for shipping to Hope Aero.

The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
Aero in 2006. It was completely disassembled, cleaned and inspected. The
external parts were very badly corroded, with most damage being to the aluminum
blades and dome shells, and rust to the steel hub and tail shaft, which was
likely caused by the outside storage environment. The inside parts were in
relatively good condition. It was established, that both technical and visual
effort would be required to restore the propeller to an acceptable display
condition.

The restoration work commenced with the blades, the alcohol de-icer feed shoes,
troughs and paint being removed. The corrosion of the blade surfaces was such
that removal by grinding was abandoned, as the inner half of the blades are
shot peened and grinding would remove this and the authenticity. It was decided
to media blast the complete airfoil surfaces with an aluminum oxide media, thus
removing the surface corrosion but leaving a good surface for paint to be
adhered to; an aviation grade grey paint was applied with yellow stripes at the
tips. New alcohol feed shoes were installed, however the troughs, which direct
the alcohol to the shoes, were not available. As the troughs are hidden inside
the spinner, it was determined that they could be omitted.
The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
tubes, and dome retaining nut, were extremely rusty. They were glass bead
blasted to remove the rust; this did not remove the heavy pitting but did leave
a surface that could be cadmium plated.  After plating, a clear coat material
was sprayed on these parts to improve the resistance to rust. The internal
surfaces were in good condition, and they were well oiled.

A replacement dome shell in relatively good condition was found, requiring only
to be sanded, anodized, and be reinstalled with new seals.

The propeller was then completely assembled to confirm that all the refurbished
parts fit together. The blades were then removed to facilitate being shipping
back to the museum.

All of the four propellers were  found to be in similar condition, and all were
restored in a similar manner.

>>>>

\title{APS 42 Radar -- Another Piece of the North Star Puzzle}

%\author{Based on Bruce Gemmill's notes; reviewed by Garry Dupont, and with editing assistance from Richard Lodge}

@@

% \paragraph*{Engine 3}

\textit{The article below was written by our members John Makadi and Chris McGuffin. Chris and John played key roles in the acquisition of the APS 42 Radar for the North Star aircraft. In 2019 John was approached by Rejean Demers, CASM's project manager for the North Star restoration to see if he could locate the radar unit which was missing from the aircraft as explained below. John was successful and as a result in 2001  Rej asked Chris if  our association  could arrange for the acquisition of the unit. John and Chris coordinated its acquisition and what follows is a very fascinating story about what may be the only unit of its kind still in existence. The authors have been very modest about their contributions in dealing with a very challenging process made more difficult by the pandemic Covid-19.}

The North Star was the RCAF’s first strategic lift platform. It was dispatched across the country, and bridged the Atlantic and Pacific in service to Canada. In the late 1950s, the RCAF began upgrading the fleet with the addition of the APS-42 military navigation/weather radar. This radar was still sensitive equipment in 1966 when North Star 17515 was donated to the Canada Aviation and Space Museum. The radar was stripped out of the airframe along with the radios and other sensitive equipment.

Project North Star is frequently described as an effort to return 17515 to the condition she was in on her last day of service with the RCAF. The hunt was on. In 2019 Project North Star Association of Canada began searching for an APS-42 radar. Initial results were discouraging. US aviation museums with USAF aircraft had no components to offer. Further research of US military logistics documents revealed that US Department of Defense disposal instructions for APS-42 radar was to “Destroy” them due to their sensitive nature at the time.

\begin{figure}[H]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{radar_the_unit.png}
   %caption of the figure 
   \caption*{\small \em Radar -- The Unit.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:img1}
\end{figure}

We were disheartened by that discovery but certainly not defeated. Over the years, PNSAC volunteers have been resourceful about rescuing parts and assemblies. The search continued. In the spring of 2021, while continuing on-line research John stumbled across an eBay listing for an APS-42 transmitter out of Cross Timbers, Missouri (Ozarks). The seller, a military communications enthusiast, acquired the radar in a lot purchase of US Government surplus equipment. After negotiating an acceptable price for the radar, we just needed to get it to CASM.

\begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{radar_the_aps_42.png}
    %caption of the figure 
    \caption*{\small \em Radar -- The APS-42.}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img2}
 \end{figure}

Shipping freight is usually pretty straight forward. In the summer of 2021, there were “unusual” circumstances. Transportation was a bottle neck to many industries and costs rose across the carriers. We were unable to find a freight company willing to collect the radar in the seller’s remote location. Our seller was willing to deliver to Springfield, Missouri (for a fee) but would not provide a crate. Furthermore, the freight companies that offered custom crating services were prohibitively high. Then we made contact with the “Air and Military Museum of the Ozarks” (AMMO) – a small volunteer-based museum in Springfield. The volunteers at AMMO were sympathetic to our cause and enthusiastic to help. They agreed to receive our radar from the seller, build a custom crate and transfer it to the freight carrier. We were grateful to pay them a small honorarium which wasn’t much more than the cost of lumber for the crate. This was a wonderful example of museums cooperating across borders to preserve aviation heritage. AMMO ended up hosting the crated radar for several weeks. The hurdles of NAFTA attestations and the export of US military technology took additional bureaucratic kung-fu but the radar arrived safely in Ottawa on 10 June, 2021.

 \begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{radar_opening_the_unit.png}
    %caption of the figure 
    \caption*{\small \em Radar -- Opening the Unit.}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img3}
 \end{figure}

After 17 months of social distancing, Project North Star volunteers and CASM staff gathered in the parking lot to celebrate the arrival of a missing radar. Customs seals were removed and a wood crate was opened to reveal our beautifully preserved APS-42. We look forward to seeing this new artifact in the nose wheel well of 17515.

% \begin{figure}[H]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-1.png}
%    %caption of the figure 
%    \caption*{\small \em North Star being moved inside hangar 11 August 2005 .}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-2.png}
%    %caption of the figure 
%    \caption*{\small \em  Mike Hope and Bruce Gemmill assembled Prop 1ready to
% install on Engine 1 October 26, 2008.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.55]{four-merlins-3.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 1 assembled in the Conservation shop Feb 2,
% 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.55]{four-merlins-4.png}
%    %caption of the figure 
%    \caption*{\small \em Firewall ready, auxiliary gearbox installed Jan 7,
% 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.55]{four-merlins-5.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 1 complete at the Restoration Shop Feb 9, 2010
% .}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.45]{four-merlins-6.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 1 installed Feb 24, 2010. Shown top left to
% bottom right, engine
% supervisor Ted Devey,  Bill Tate, Ron Lemieux, Giuseppe Zanetti, John Corby,
% Garnet Chapman, Volunteer Project Manager Jim Riddoch, John Tasseron, Bruce
% Gemmill, Museum Project Coordinator Mike Irvin, Charles Baril, and then
% Association President Tim Timmins.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \paragraph*{Engine 2}\

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-8.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 2 with cowl panels removed, April 6, 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-9.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 2 removed April 14, 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-10.png}
%    %caption of the figure 
%    \caption*{\small \em Stan Rideout working to reassemble Engine 2 in engine
% shop, Oct 7, 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-11.png}
%    %caption of the figure 
%    \caption*{\small \em Austin (Tim) Timmins.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-12.png}
%    %caption of the figure 
%    \caption*{\small \em Jim Riddoch and Bill Tate fill the radiators with
% inhibiting oil.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.7]{four-merlins-13.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 2 on rotating assembly stand.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.45]{four-merlins-14.png}
%    %caption of the figure 
%    \caption*{\small \em  Engine Nr 2 being re-assembled in the engine shop.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-15.png}
%    %caption of the figure 
%   % \caption*{\small \em .}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

### 2022-04-08 pnsac/newsletter/article_PresReport.tex
First set of changes for 18-1 issue, Apr.2022.

\title{Notes from the President}
\author{Richard Lodge}

@@

As always, before I start to write the Notes from the President, I reread my
previous Notes to remind myself of what I had previously said in order to set
the record straight if necessary. As I write this at the beginning of
September, I am now aware of the impact of the COVID-19 pandemic on the
restoration schedule for the North Star and the operations of your
Association.

Since March, restoration work on the aircraft has been suspended and the
activities of the Association have been limited. We have, however, been keeping
in touch with each other through the weekly Zoom meeting organized by the
Museum's volunteer co-coordinator Cédric St.-Amour. This virtual meeting takes
place every Thursday at 10 AM EST; it is open to all our volunteers and any
other PNSAC members who would like to join in.

As previously reported, the North Star project reached a major milestone with
the completion of the restoration of Engine \#4 which was moved into the
reserve hangar on a cold December 2019 morning.

Since we are unable to physically meet and there has been no restoration work
carried out, we felt that a special issue of the NSTAR Chronicle devoted
entirely to the 15 year restoration of all four of the Merlin engines would be
of interest to members.

Understandably, with a project which has taken many years. the volunteers
working on the engines have changed over time. Some have retired or stopped
volunteering and sadly, some have died. Earlier this year we asked Ted Devey to
contribute a piece but before writing about of his work, he passed away. Ted
was involved with restoring all the engines and for several of the early years
was the volunteer leading the restoration team. Ted was a great raconteur and
our oldest volunteer at 93 years old. His recollections would have added
considerably to the interest of this special issue of the NSTAR
Chronicle.

After much consideration, the Museum has decided that volunteering cannot
restart before January 2021. Until then we will continue with our regular Zoom
meetings and will also organize a virtual Annual General Meeting for the
Association. We will also continue to provide updates on the situation and much
will depend upon how the different levels of government handle the ongoing
recovery from the pandemic.

There is an old saying "That it is an ill wind that blows nobody any good."
This is very true even of this pandemic. One of the very pleasant results is
having conversations with PNSAC members who don' t reside in the National
Capital area and have sent membership renewals by mail. Where a renewal has
been received and confirmation has been delayed as a result of the pandemic, I
try to call a member on the telephone to explain what is happened and that the
renewal has not been lost. As a result I have had many interesting phone calls
with people whom I have never spoken to before. A definite bonus for me.

>>>>

\title{Notes from the President}
\author{Chris McGuffin}

@@

I showed up for the freebie, got an education and made some friends. That was my start in aviation. In 1985 I signed up for Air Cadets for something to do. One could say my mother made me. Certainly, she encouraged. The Cadets played sports and hosted camps; for free. I had no idea about the flight scholarships and other opportunities. Certainly, I was enamored with aviation. I built model airplanes and made a wind tunnel for a science fair. That wasn't much different from my sister's engagement with visual art or my brother's fascination with insects. My real passion was fixing things, improving them or making new things. My mum probably just wanted me to get out of the basement and socialize.

Aviation was exciting but I never considered earning a living from it. I studied electrical engineering at the Royal Military College. I thought a little military service would be a precursor to advancement in the civilian tech industry. Instead of Nortel I ended up serving 28 years with the Canadian Forces. Many of my friends wore the blue uniform. I wore army green or tan depending on the season. When I flew it was usually to participate in a meeting, attend training or visit a forward operating base. Still, there can be a persistence to childhood interest.

My interest in wood and metal working matured as a hobby. I have a particularly thoughtful wife who accommodated a larger work space with every move. Nonetheless, a residential workshop can be a solitary and confined space. When I retired from military service in 2017 it was time to look for different ways to pursue my interests. It would be fun to learn from new people and with tools I didn't have access to at home. I could not imagine a better opportunity than an aircraft restoration.

My timing lined up with Project North Star, a comprehensive restoration effort: metal fabrication, engine repair, refinishing, reupholstery, overhauling mechanical systems, woodworking and more. Some volunteers contribute decades of skill and expertise while others offer enthusiasm and commitment to learn. Everyone benefits from cross training and camaraderie. Progress is important but not more than the historical value and preservation of the artefact. Each step is supervised by our project manager, an aircraft maintenance engineer employed by the Museum.

Initially, I was assigned to reproduce and restore the wood paneling that lined the interior of our North Star. Occasionally I helped with other assignments. I watched as replacement parts were fabricated, engine \#4 was reassembled and bird nests were pulled out of the tail section. In 2018 I began publishing progress and event updates on the Project Facebook and Instagram pages. Many people enjoy seeing airplanes up close. Airshows are popular for good reasons. I certainly look forward to the time I spend at the aviation museum, particularly the areas normally hidden from public view. Taking a few photos to share with our social media audience was a natural fit for me. Those updates have been few during the pandemic but they will resume imminently.

2022 is going to be a year of renewal in many ways. In January I assumed the presidency of the Association. Our previous president, Rodger Lodge, held the position for nearly 12 years. Roger has been involved with the Association for 20 years and I am thankful that he has agreed to stay on the Board of Directors. The Canada Aviation and Space Museum is actively renovating the conservation workshop. Plans for a return of volunteers to the Museum are being developed and restoration activity should resume in the months ahead. I am grateful for the encouragement of our supporters and the tenacity of our volunteers. We'll be back to the restoration effort soon.
